### 2023-09-16 review/icfp23/response.md

Thank you for your thoughtful and detailed comments! We believe that the issues raised can be
addressed with only small revisions.

# Extensions (RA, RB, RD)

The reviews suggested extensions of the marked lambda calculus to handle polymorphism (RA, RD),
linear types (RB), dependent types (RB), and subtyping (RD). Our intention was to present a minimal
framework that, like the STLC, is intended to establish guiding metatheory–most importantly,
totality–in support of conservative extensions in many subsequent directions. We dedicate Sec. 2.2
and 2.3 to two small examples of such extensions, emphasizing how totality guides us. We do
similarly consider System-F-style polymorphism in Sec. 2.5, albeit less formally for space reasons;
we would be happy to use an extra page in the final version to include the straightforward rules for
“marked System F” if requested.

We conclude Sec. 2.5 with a perhaps too brief discussion of more sophisticated extensions, e.g.
implicit type application. We would be happy to more explicitly lay out the general approach, which
is still to first define a bidirectional + gradual version of the feature, then use totality as a
guide. We will cite relevant work on gradual implicit type application [1], gradual dependent types
[2], and gradual subtyping [3] in the final version of this section; we aren’t aware of a gradual
linear type system, though there are gradual session type systems [4]. However, it is beyond the
scope of this initial paper to define marked dialects of these systems, which each comprise dozens
of typing rules.

Our implementation does scale beyond the calculus to those features found in ML-family languages
like Elm, e.g. full pattern matching.

We agree that marked dialects of advanced type systems might require a designer to experiment with
different localization heuristics when errors are due to a confluence of factors (e.g. as in Sec.
2.1.6 with conditionals). Our contribution is to provide a framework for clearly specifying these
design decisions, rather than leaving them implementation-defined.

# Mechanizing constraint solving / Sec. 4 contributions (RB, RC, RD)

While there are several mechanizations of unification-based type inference (e.g. [5]), the
metatheory of these systems (e.g. soundness, principality) is relevant only when the constraint set
is consistent. Our focus in Sec. 4 is on inconsistent constraints: suggested fillings can be
inconsistent with the constraints. There is no critical metatheorem, like totality, that must guide
this development; the approach in Sec. 4 is instead fundamentally rooted in the heuristic that even
inconsistent constraints likely contain information about user intent. The key technical challenge
is therefore in efficiently computing partially consistent suggestions, so we intentionally favored
an algorithmic presentation in Sec. 4.4.

We also intentionally used simple, well-studied techniques, like union find, because the real
contribution of Sec. 4 is the system architecture: we demonstrate that one can dramatically clean up
type error localization by using constraint solving only for localizing erroneous type holes,
layering it atop a bidirectional + gradual core where localizing erroneous expressions is more
systematic. Previously, these approaches were seen as alternatives. We agree that Sec. 4 could use a
clearer “punchline” along these lines (RC, RD) and general clarity improvements (many suggested by
the reviewers).

# Related work: interleaving (RA)

We agree that the approach taken in contemporary constraint-based type inference implementations,
cited by RA [6, 7], of interleaving constraint generation and solving could inform efforts to
interleave marking (Sec. 2) and constraint solving (Sec. 4). However, because Sec. 2 can be adopted
on its own, we believe that introducing these topics separately, based on more elementary and
widely-understood foundations (e.g. STLC, HM) is expositionally preferable. We will discuss.

# Empirical evaluation (RB, RC)

The primary contribution of Sec. 2 is a metatheoretic framework, so our primary evaluation of this
contribution is the mechanized metatheory. While it would be interesting for future work to, for
example, empirically evaluate different approaches to localizing inconsistencies between conditional
branches, we make no specific usability claims and formalize each possible approach.

The primary contribution of Sec. 4 is a novel architecture, described above, that simplifies and
systematizes a previously complex and ad hoc task: constraint inconsistency error localization.
Certainly it would be interesting to empirically evaluate the “code completion” affordance we
present that allows quick toggling between partially consistent type hole fillings, but we again
make no specific usability claims about this particular interface. It is optional and can be toggled
off without any loss of liveness elsewhere in the system, which relies only on the localization
mechanisms detailed in Sec. 2.

We have deployed Hazel in the classroom recently, so we hope to observe how students interact with
all aspects of the relevant affordances, including localization choices, error messages, and code
completions, over the coming years, but we believe that the metatheoretic and architectural
contributions of this paper open up an interesting and novel region in the design space on their
own. We will be careful to make clear that there are no specific HCI claims related to the UI
screenshots or specific error message text in this paper.

# What is “gradual recovery”? (RC, RD)

We will clarify in the introduction that by “gradual recovery” we refer to our use of gradual typing
to both recover from missing type information (pg. 3, Sec. 2) and to provide a site for
interactively recovering from conflicting constraints (pg. 4, Sec. 4).

# Motivation for Sec 3 (RA, RD)

Section 3 is a short section pointing out that marking allows us to resolve a significant problem
related to undefined behavior in the Hazelnut edit action calculus. We introduce the judgemental
forms but indeed, a deep understanding of this section would require familiarity with Hazelnut. We
can note that this section can be safely skipped for readers uninterested in formal models of
structure editing. (It can also be omitted from the final version if requested.)

# Additional Responses

We respond to some more specific questions brought up by each reviewer below.

## RA

    A good part of the early presentation is a bit repetitive in nature, and could be summarized by
    stating that valid unmarked typing derivations are left as-is in the marked word. Is that
    correct?

Yes, this is formalized with Theorem 2.3, part 1. We can mention this sooner.

    Why aren't destructuring bindings implemented like this at all? The proposed way seems much more
    complicated, more expansive, and it's unclear to me why it would provide better error messages.

Sec. 2 is intended as a standalone system that does not incorporate constraint solving, so this was
not an option. We included this in the paper because a strictly bidirectional approach is indeed
rather subtle and we hadn’t previously seen this issue brought up, e.g. in Dunfield’s 2018 review
paper. The approach we specify ensures that information from both the expression and pattern is
unified, while taking pattern type annotations as “ground truth” for localizing errors in the
expression. Sec. 4’s approach would generate the constraints as RA describes.

    The use of ?u for unification variables confuses me a lot. It seems to mix up unification
    variables with the gradual unknown, which seems completely wrong to me. A gradual unknown is any
    type (possibly several different ones during the execution), a unification variable is a single
    type that isn't determined yet. Am I missing something motivating your usage? (RA)

We consider missing type annotations type holes for the purposes of inference, i.e. we presume the
user intends to fill type holes with a known type. To expose gradual typing for intentional use, we
could introduce a “dyn” type that would fill a type hole and behave identically except that it would
not be subject to type hole filling. (This is related to the approach explored by Garcia and Cimini,
which was mentioned by RD.)

    If I remember correctly, Elm also uses this graph to diagnose type errors, by recognizing
    specific patterns in the graph of types and/or constraints. Is that something you do as well,
    and can you develop a bit on that? (RA)

We take the opposite approach, as we described above. Briefly, we do not attempt to try to localize
constraint solving failures to expressions, leaving that to the marking system. This avoids the need
to define this style of ad hoc system – type inference variables that cannot be solved can be
systematically localized to type holes, and the user can interactively choose from partial fillings
to see where the problem would be localized in a second step.

# RB

    can I essentially think of this as type checking, but that it will always produce a type where
    there may nevertheless be marks to indicate the location of type errors? And an unmarked
    expression is not necessarily well typed but is well formed?

Yes, precisely!

# RC

    Why are your rules algorithmic? (RC)

The rules are incidentally algorithmic from our perspective – bidirectional typing is useful because
it specifies a static “direction” that information flows, which thereby systematizes error
localization. In addition, we want to ensure unicity, i.e. that there are not multiple possible
markings, because our goal is to formalize what would normally be an implementation decision.

# RD

    Could you clarify what would happen if a conditional would be subsumable? (RD)

This would violate unicity (i.e. make marking non-deterministic) in the case where a conditional
expression with inconsistent branches appeared in an analytic position, e.g.

```
let x : Num = tt ? ff : 2
```

The analysis rule would mark the error on the “then” branch, while the synthesis rule would mark the
error on the conditional as a whole (or arbitrarily on one of the branches if the alternative rules
are used).

## Related Work (RD)

Thank you for bringing up some more distantly related work on type inference, which we respond to
below. We will include this in the final related work section.

    Garcia and Cimini's POPL 2015 paper on type inference for gradually-typed languages, which
    improves substantially on the prior work of Siek and Vachharajani.

Indeed, this paper is a cleaner presentation of the approach explored initially by Siek and
Vachharajani. It distinguishes explicit unknowns from types that must be inferred, whereas we
consider type holes as types that need to be inferred (and also allow other internally generated
unknowns to be inferred, with full provenance). We could add a “dyn” type to allow users to mark
that they intend the gradual unknown semantics and do not want a solution to be inferred (see
response to RA above) which would be consistent with Garcia and Cimini. This paper does not consider
the key problem we focused on, which is generating partial solutions, but we agree that this paper
should be discussed in more detail alongside Siek and Vachharajani in the final version.

    The second is Vazoy et al's OOPSLA 2018 paper on gradual liquid type inference, which features a
    similar treatment of holes and interactive exploration of possible solutions (in that case, for
    unknown refinements, but the underlying concepts are very close).

We agree that the liquid type exploration outlined in this paper is similar in spirit to our
approach in that it allows for interactive exploration of conflicting refinement type constraints,
localized to holes in a refinement type specification. Marking is not detailed in this paper, and
the technical approach to synthesizing refinements is quite different, but we will certainly mention
that the exploratory interface we propose has connections to this prior work!

    Also, Lennon-Bertrand et al.'s TOPLAS 2022 paper on Gradual CIC includes an integration of
    bidirectional typing and gradual typing.

We mention this at the beginning of the author response – we agree that this system would be a
suitable basis for creating a marked CIC and look forward to future explorations in this direction!

[1] Xie et al, 2019 (based on Dunfield and Krishanswami, 2013)

[2] Lennon-Bertrand et al, 2022

[3] Siek and Taha, 2007

[4] Igarashi et al, 2017

[5] Paulson, 1993 (and several other subsequent efforts, which we will cite as well)

[6] Odersky et al, 1999

[7] Pottier, 2014

### 2023-09-16 review/popl24/response.md

Thank you for your thoughtful and detailed comments. We address questions brought up by the reviewers below. The issues raised can be addressed with revisions that will correct the typos and small expositional clarity issues identified by the reviewers and expand on the discussion of the design space, as suggested by the reviewers.

# Review A

> In the rules you have chosen τ₁ from the environment. I would argue that τ is the more natural
> choice from the programmer's point of view (that's what they wrote!) and ? is the more principled
> one. At the very least, you need to acknowledge that there is a choice to make here.

This is actually a typo—indeed, we agree that it is more natural to choose τ (and in fact this is
the choice found in the Agda mechanization)! We will fix this in the final version and, in general,
more clearly highlight these spots where language designers must make such decisions.

> It's not clear at all why the first two premises [of USLetPat] (involving τp) are needed at all,
> since no information flows out of them.

Yes, those two premises are unnecessary. They were included as, following the marking logic from
ISLetPat, we must ultimately analyze the definition against the pattern type to properly attribute
any inconsistency error to the definition, given that we want any annotations in the pattern to be
canonical. However, for USLetPat, the analysis of the pattern against the definition type suffices
to ensure consistency, so the second and hence the first premise are redundant, and will be removed
from the final version, where we will clarify this point.

> Is the correspondence between this approach to polymorphism (using the gradual unknown type for
> unification variables) and standard Hindley-Milner known? Are there any systems implementing this
> currently?

Garcia and Cimini (2015) have previously developed the connection between polymorphism, gradual
typing, and constraint-based (Hindley-Milner) type inference. We discussed this paper in the Related
Work section, but will also mention it alongside Siek and Vachharajani (2008) in Section 4.4 in the
final version (and expand our discussion of it in the Related Work section).

> In section 3.1 you write that you have implemented marking for algebraic datatypes and explicit
> polymorphism, but there are no examples for this in the supplied Hazel build, and I can't figure out
> what the syntax is.

These features were in different branches of Hazel at submission time, so they did not end up in the
build provided in the submitted supplement. They will be merged in the final artifact. We will also
include an Examples tab in the UI which will demonstrate the relevant syntax.

# Review B

> Lines 644-645 say that "[i]f the pattern a suggests that the conditional is in synthetic position,
> it will be marked with an inconsistent branches error mark following our development above. If
> instead it is in analytic position against the unknown type, no mark will be produced." Is this
> desirable? I would expect to get a type error pointing out inconsistent branch types regardless of
> how the expression is type checked.

The gradually typed interpretation of type holes as the unknown (i.e. dynamic) type suggests
allowing each branch to be checked independently (as only one of the branches will be taken at
run-time). It is essentially as if each branch was independently annotated with a ? type ascription,
which would cause both branches to become consistently typed. The inconsistency between branches
would still be caught by the type hole inference mechanism described in Section 4, localizing the
problem to the ? itself. We will clarify this in the final version.

> Have you thought about how this (e.g., Section 2.1.1) would work with operator overloading (e.g.
> if you add a float type and overload + to operate on both ints and floats?)

Overloading would require changing one or both operand modes to synthetic (depending on how operator
overloading is defined—the left operand is often arbitrarily privileged in languages with this
feature) so that the overloading can be resolved during typing/marking. If an operator is not
defined given the synthesized type of the operand(s), the error might be marked on the operator
itself. Type classes (a.k.a. traits / implicits) present a symmetric alternative to operator
overloading where the operands can remain in analytic position. We will briefly discuss this
situation in the final version.

> I accept that producing error messages was not the focus of the paper, but are there more general
> lessons you can draw on error localization (e.g., for other languages/paradigms, or if one wants to
> do better error localization but without buying fully into the mechanisms of typed holes)?

The marked lambda calculus relies in some instances on having an unknown type à la gradual typing
(which has been studied and applied across many language paradigms, including imperative / OO
languages). However, we will note more clearly that empty expression holes are not integral to the
core calculus and recipe—they are included to illustrate how the marked lambda calculus fits into
the larger problem of modeling incomplete program states and to discuss polymorphic generalization
in Section 4.4, but could be removed entirely from the marked lambda calculus without issue. As
such, the techniques described in the paper should be relevant to a wide variety of language
designs, including those where hole-driven development is not part of the user workflow.

# Review C

> line 288: I was expecting some kind of "Localisation" principle about pushing errors to
> a maximally local position. Is that not possible? It seems related to "the neutral approach."

We did not consider the problem of measuring locality but we note that the typing rules for marks
ensures that the marks are placed precisely at locations where inconsistencies exist, and therefore
less local marks (e.g. on parent expressions) would violate the metatheory. Bidirectional typing
specifies a flow of typing information that ultimately justifies the localization decisions in the
marked lambda calculus; the justification for the type flow of bidirectional typing is discussed by
other authors, notably Dunfield and Krishnaswami (2019).

> line 442: what would break if you synthesized the body in the premise of the rule rather than
checking against? Anything? Is there a guiding principle around this?

It is indeed possible to instead synthesize the body here, but we chose analysis because the overall
lambda is in analytic position so, were the expected type an arrow type, the body would be in
analytic position and therefore we retain this mode using the unknown type. The design principle
here is perhaps some notion of “conservation of mode”, which we can discuss in the final version,
but it is not critical here.

> line 452: In the premise of MKALAM3, why is x typed \tau_1, as opposed to \tau or "?"? What is the
> guiding principle at play here?

As mentioned above in response to a similar question in Review A, this is actually a typo, and
indeed we believe it more natural to choose τ (and this choice is the one found in the Agda
mechanization). This will be corrected in the final version.

> line 458: In the USAp rule, is it necessary for e2 to be checked against \tau_1? Is it plausible
> to instead synthesize the argument type and then "half-check/half-synthesize" the function? It would
> be nice to understand what the boundaries of this design are.

We are not entirely sure what is meant by “half-check/half-synthesize”, but our development of the
unmarked typing rules follow the typical formulation of function application in bidirectional
typing. Dunfield and Krishnaswami (2019) discuss other “modes” for function application where
information could flow in other directions, and if these alternatives were chosen, error
localization would need to be updated accordingly.

> line 531: Why must localization be deterministic? Given the number of design decisions involved,
> is it plausible to defer them or choose in some delayed manner, or by using more program context
> to decide?

The algorithmic and deterministic nature of the localization calculus makes it straightforward to
implement as part of a real type checker. Ultimately, how errors are localized depends heavily on
the formulation of the original type system. However, it may be possible to involve the user or
other heuristic mechanisms to resolve ambiguity rather than specifying a decision
deterministically—we explore this in concept in Section 4 for type inference errors in particular,
and will mention it more broadly as a potential direction for future research.

> Section 2.3 seems to reuse the ? type to model phenomena that should make sense in a language that
> is bidirectionally-typed, has pattern matching, and has no notion of gradual typing whatsoever.
> I think that this would be a much better contribution if it started with explicating a purely static
> version of bidirectional pattern matching and then demonstrated that the marking recipe transfers
> over to this purely-static version. I expect that such a system would need some way of accounting
> for partial analysis and synthesis, but without using gradual typing (albeit something related but
> static).

The type ?⇒ for switching to synthesis does indeed not require gradual typing and is merely
a marker for switching to synthesis because we are using analysis to propagate information from the
pattern. It never needs to participate in a type consistency relation (via subsumption) because the
UASynSwitch rule would always apply there. It need not look like an unknown type, i.e. using ? may
be misleading notation, so we will clarify this as suggested in the final version.

> line 678: the USLETPAT rule seems to depend on all expressions being able to synthesize, which is
> not the case in all bidirectional systems (e.g. the common treatment of unannotated lambdas). Can
> this be generalized?

Yes, we do assume that successful analysis implies synthesis, but a more general treatment here
would choose to use the pattern’s synthesized type to analyze the bound expression in situations
where no type can be synthesized. We will describe this design dimension in the final version.

> line 730: "these marked types operate identically..." why is that the "right" thing to do? What is
> the guiding principle (aside from it seeming to work)?

In the same way that unbound variables and other error-marked expressions in our core development
synthesized the unknown type to allow type-checking to continue, so too do unbound type variables
behave like the unknown type.

> line 816: what is the box superscript?

This is mark erasure, as defined in Figure 3.

> line 828: I find it strange to call bidirectional typing "local type inference" since local type
> inference (at least in Pierce and Turner) was a combination of bidirectional typing and constraint
> solving.

We will make this distinction more clear in the final version.

> line 884: What is the red angle bracket? I didn't see it explained.

The editor cursor in Hazel is drawn as an angle bracket at expression boundaries to convey the
“hexagonal block structure” of Hazel expressions (see Moon et al, VL/HCC 2023 for details – we will
clarify in the final version).

> line 963: "without needing to introduce explicit constraints" I did not understand this, since the
> rules seem to be adding explicit constraints.

We will clarify that the purpose of the matched arrow provenances here is to avoid generating fresh
type inference variables each time a matched arrow type is invoked, which would lead to a blow up in
the number of inference variables that must necessarily be equal to one another.

> line 969: "A marked expression has already been deemed erroneous" but why is it the expression,
> rather than the boundary between outside and inside?

The bidirectional flow in Section 2 determines expression error localization, and indeed it is
typically due to inconsistencies between the type information “outside” and “inside” the marked
expression that causes issues.

> line 982: "a standard union-find based unification algorithm" but this algorithm must account for
> consistency, right?

Yes, the algorithm favors more precise types as previously described e.g. by Siek and Vahharajani
(2008) and Garcia and Cimini (2015). We will make this point more clear.

### 2023-10-24 let.tex
Updates from Overleaf
USLetPat and related adjustments

To ease the use of products and other datatypes, many languages feature destructuring bindings.
In a typed setting, we may want to granularly add type annotations in patterns as well.
As it turns out, this combination of features is surprisingly tricky to get right!
Let us add let expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a variable pattern, and $\PPair{\PMV_1}{\PMV_2}$ for a pair:
%
\[\begin{array}{rrcl}
  \EMName  & \EMV  & \Coloneqq & \cdots \mid \ELet{\PMV}{\EMV}{\EMV} \\
  \PMName  & \PMV  & \Coloneqq & \PWild \mid \PVar{x} \mid \PPair{\PMV}{\PMV} \mid \PAsc{\PMV}{\TMV}
\end{array}\]
%

@@

To type this, the most obvious approach is to synthesize a type from the pattern and analyze the
definition against that type. However, this may run afoul of user expectation, which might
reasonably suppose it to be equivalent to the expanded expression
$\ELet{\PVar{a}}{1}{\ELet{\PVar{b}}{2}{\EMV}}$. In the original, the pattern synthesizes the type
$\TProd{\TUnknown}{\TUnknown}$, and $\EPair{1}{2}$ is analyzed against it, hence $1$ and $2$ are
each analyzed against $\TUnknown$. In the expanded version, however, they are typed in synthetic
mode.

@@

To ensure propagation of type information between pattern, definition, and body, the destructuring
let requires some extra machinery.
\[%
  \judgment{
    \ctxSynPatU{\ctx}{\PMV}{\TMV_p} \\
    \ctxAnaTypeU{\ctx}{\EMV_{1}}{\TMV_p} \\
    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\\\
    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
    \ctxSynTypeU{\ctx'}{\EMV_{2}}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\TMV_2}
  }{USLetPat}
\]%
First, to ensure that the pattern and definition types are consistent, the pattern synthesizes a
type, against which the definition is analyzed. Then, the definition synthesizes a type, against
which the pattern is analyzed. This analysis is guaranteed to succeed; we need the context it
produces for use in the synthesis of the body's type. The analytic rule is similar, with the final
premise and conclusion changed to analysis.

The marking rule is directly analogous and utilizes the marking of patterns, which we omit---they
may be derived in the same way as those for expressions and are governed by similar metatheorems (again, totality guides the derivation).
In the analytic cases, we introduce error marks that parallel to $\ECInconType{\ECMV}$ and 
$\ECAnaNonMatchedProd{\ECMV}$.
\[%
  \judgment{
    \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\TMV_p} \\
    \ctxAnaFixedInto{\ctx}{\EMV_{1}}{\ECMV_{1}}{\TMV_{p}} \\
    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\\\
    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
    \ctxSynFixedInto{\ctx'}{\EMV_{2}}{\ECMV_{2}}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\ELet{\PCMV}{\ECMV_{1}}{\ECMV_{2}}}{\TMV_2}
  }{ISLetPat}
\]%

>>>>

To ease the use of products and other datatypes, many languages feature destructuring bindings.
In a typed setting, we may want to granularly add type annotations in patterns as well.
In a bidirectional setting, as it turns out, this combination of features is surprisingly tricky to get right!
Let us add let expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a variable pattern, and $\PPair{\PMV_1}{\PMV_2}$ for a pair:
%
\[\begin{array}{rrcl}
  \EMName  & \EMV  & \Coloneqq & \cdots \mid \ELet{\PMV}{\EMV}{\EMV} \\
  \PMName  & \PMV  & \Coloneqq & \PWild \mid \PVar{x} \mid \PPair{\PMV}{\PMV} \mid \PAsc{\PMV}{\TMV}
\end{array}\]
%

@@

To type this, one approach is to synthesize a type from the pattern and analyze the
definition against that type. However, this may run afoul of user expectation, which might
reasonably suppose it to be equivalent to the expanded expression
$\ELet{\PVar{a}}{1}{\ELet{\PVar{b}}{2}{\EMV}}$. In the original, the pattern synthesizes the type
$\TProd{\TUnknown}{\TUnknown}$, and $\EPair{1}{2}$ is analyzed against it, hence $1$ and $2$ are
each analyzed against $\TUnknown$. In the expanded version, however, they are typed in synthetic
mode.

@@

Let synthesis is straightforward: we synthesize the definition's type, then analyze the pattern against it:

\[%
  \judgment{
    %\ctxSynPatU{\ctx}{\PMV}{\TMV_p} \\
    %\ctxAnaTypeU{\ctx}{\EMV_{1}}{\TMV_p} \\
    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\
    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
    \ctxSynTypeU{\ctx'}{\EMV_{2}}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\TMV_2}
  }{USLetPat}
\]%

The analytic rule is similar, with the final
premise and conclusion changed to analysis.
\\
Attempting an analogous marking rule runs into trouble. We want any type annotations in the pattern to be canonical, which means analyzing the definition to attribute to it any inconsistencies. But we still must analyze the pattern, incorporating the definition's type to produce the context needed by the body. So we do a 'round trip': Analyze the definition against the pattern's type, and then the pattern against the definition's. Since the first analysis establishes consistency, this second analysis is guaranteed to succeed; we only care about the context produced:

\[%
  \judgment{
    \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\TMV_p} \\
    \ctxAnaFixedInto{\ctx}{\EMV_{1}}{\ECMV_{1}}{\TMV_{p}} \\
    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\\\
    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
    \ctxSynFixedInto{\ctx'}{\EMV_{2}}{\ECMV_{2}}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\ELet{\PCMV}{\ECMV_{1}}{\ECMV_{2}}}{\TMV_2}
  }{ISLetPat}
\]%

We omit the marking of patterns---they
may be derived in the same way as those for expressions and are governed by similar metatheorems (again, totality guides the derivation).
In the analytic cases, we introduce error marks that parallel to $\ECInconType{\ECMV}$ and 
$\ECAnaNonMatchedProd{\ECMV}$.
